### 2014-11-18 papers/paper1.tex
First commit

\chapter*{A Brief History of Computer Games}

  \section*{Game Changes the past fifty years}

  \begin{itemize}
    \item Changes in hardware for playing games
    \item Changes in interaction devices
    \item Changes in the software tools available
    \item Changes in game business
    \item Changes in the demographics of the players
    \item Diversification
    \item Changes in the design of games
  \end{itemize}

  \section*{Game History}

  \subsection*{1950-1959}
    \begin{itemize}
      \item First computer game {\bf OXO}, a version of {\bf tic-tac-toe} (1952)
      \item People consider the first interactive computer game to be {\bf Tennis for two} (1958). Effects of gravity.
      \item The first game developers didnt understand the potetial of games because of the hardware/equpment needed.
    \end{itemize}

  \subsection*{1960-1969}
    \begin{itemize}
      \item {\bf Spacewar} (1961)
      \item Companies started to consider the commercial explotation of computer games.
      \item 1966 Sega released the {\bf arcade game Periscope} 
    \end{itemize}

  \subsection*{1970-1979}
    \begin{itemize}
      \item The golden age for arcade games
      \item The first commercial explotation of computer games came through arcade machines.
      \item The arcade machines costed money, making it commercially feasible to exploit computer games. 
      \item The first arcade computer game {\bf computer Space} appeard in 1971, but was not a commercial success
      \item {\bf Pong} was a big commercial success created by {\bf Atari}(1972). Following commercial successes was {\bf Breakout} and {\bf Space Wars}.
      \item {\bf Space Wars} was the first game using vector graphics
      \item in 1978 color was introduced
      \item {\bf Space Invanders} and {\bf Asteroids}
      \item {\bf Death Race} (1976) led to a lot of controversy which led to its end.
      \item Research on interactive televison resulted in the {\bf Odyssy game console} (1972). Could only move some dots on the screen and needed plastic overlays to the TV to add colered playfields.
      \item {\bf Channel F} (1976) making it possible to play different games on the same system.
      \item The big step was when Atari introduced the {\bf VCS system}. The console did not sell well becuase it was expencive and the games was not impressive. It became a success when they introduces Space Invanders. This shows that it is not the hardware that counts, but the games.
      \item The hardware on the CVS was limited (1 kb of memory for program and data) and program was written in assembly. This made it hard to program interesting games. 
      Since there was limited memory, the playing fields was often symmetric. This saved data.
    \end{itemize}

  \subsection*{1980-1989}
    \begin{itemize}
      \item {\bf Pack-Man} (1980), led to coin shortage in Japan.
      \item Many interesting games was introduced in this period: {\bf Pack-Man} (1980), {\bf Donkey Kong} (1981), {\bf Mario Bros} (1983), {\bf The leged of Zelda} (1986), {\bf Final Fantasy} (1987), {\bf Prince of Persia} (1989).
      \item Because of the success of Atari, others followed, as well as companies appeard creating games for the different game consoles. 
      \item Atari failed when making the game {\bf E.T}. Many game companies went bakrupt or stepped out of the game business. Game production moved to Japan. 
      \item Another reason for {\bf the crash} was the introduction of new game computers. Cheap {\bf PC's} appeard and was suited for games because of memory, graphics and sound capabilities. {\bf Commadore 64}.
      \item Games for computers was {\bf easy to copy} because of the floppy disks or cassette tapes. 
      \item The pc games made it possible to {\bf save game progress}. 
      \item The crash of the console market made it possible for other companies to enter the market, like {\bf Nintendo NES} (1985) and {\bf Sega Master System} (1986).
      \item {\bf Super Mario Bros} on NES was a huge success. The NES was more popular, not because of its hardware, but because of its uniqeness of its games. 
      \item The {\bf D-pad} was introduced and replaced the joystick. 
      \item Nintendo introduced the first handheld gaming system {\bf Game Boy} (1989). Came bundled with {\bf Tetris} that became a huge success. For a long time Nintendo was the prime producer of handhelds. 
    \end{itemize}

  \subsection*{1990-1999}
    \begin{itemize}
      \item New game consoles was introduced {\bf Mega Drive/Genesis} and {\bf Nintendo Super NES}. They had better hardware. Special hardware for drawing sprites. Higher screen resoultions.
      \item Nintendo had {\bf Mario} as their main character, Sega introduced {\bf Sonic the Hedgedog}.
      \item There were other systems around, but Sega and Nintendo had the Majority of the market.
      \item The newcomer was in the game console was {\bf Sony} who relesed {\bf PlayStation} (1994). PlayStation had faster processor, more memory and special hardware for 3D. 
      \item Games for PlayStation were easier to programming. Resulted in higher game production. 
      \item Players wanted more complex games and it resulted in a {\bf change in the game business}. Huge game budgets became common. 
      \item PC became mature. Many great games were produced like {\bf Sim City}.
      \item PC's could stream video and music from CD. This led to a {\bf new generation of games} that relied on good integration of video and sound. 
      \item PC's had the advantages with the mouse and keyboard. 
      \item PC's had {\bf modem}. This led to the rise of the many massive multiplayer online role playing games  {\bf MMORPG}. 
      \item {\bf The nerds takes over!} The games on the PC was hard to install making the most interested nerds the only ones playing. 
      \item PC's had {\bf different hardware specs}, making it hard to program games for all the different hardware specs. 
      \item Windows 95 and the release of {\bf DirectX} (1995) abstracted away underlying hardware. 
      \item {\bf Doom} ``the first'' first-person shooter game with fake 3D. 
      \item Because of the success of the 3D like game Doom, there was an increased interest in 3D graphics cards for PC.
      \item Handhelds got a new generation, {\bf Game Boy Color} (1998). It had better spec than the older one, and it could communicate wither other devices. 
      \item {\bf Pokemon} became a important game for the Game Boy Color. Pokemon added the important collecting aspects. Nintendo introduced different colors, and players had to communicate with other players in order to collect all possible pokemons. 
    \end{itemize}

  \subsection*{2000-2009}
    \begin{itemize}
      \item Sega decided to stop the production of game consoles in 2001, but continued to make games for other game consoles. 
      \item In 2000 the {\bf PlayStation 2} was released. Better specs, excellent sound qualities, network adapter, DVD player.
      \item The DVD feature in in PS2 was responsible for a quite early sales because PS2 was cheaper than other DVD players. The problem was that people that bought PS2 as a DVD player did not buy any games. PS2 had backwards compatible with PS1. PS2 games was hard to program, but a huge amout of popular games was lauched and the PS2 became popular. 
      \item Nintendo followed in 2000 with its {\bf GameCube}. They did not focus on specs and was weaker than its competition, but a huge amount of great games (casual games) and the low prize made it popular. 
      \item In 2001 Microsoft entered the market with {\bf Xbox}. It is rumored that Microsoft started their own console after Sony refused to use DirectX. The Xbox was a powerful machine that basicly was a PC in a console box. One of the top titles was {\bf Halo}. Good in the {\bf online domain}.
      \item Xbox was lauched late, and had a hard competion with the available SP2 with a lot of available games. 
      \item Creating games became for new consoles became a more complicated and expensice. Players wanted better graphics, movies, sound tracks, and increased playing time. {\bf Game budget increased} because they needed bigger teams with programmers, artists, and much more. 
      \item PC's was getting better specs and the hardcore gamers prefered the PC over the game consoles. This araised the problem to fit the different specs on the different computers. 
      \item Another problem with PC games was that they was easily cracked and it became harder to make money on PC games. 
      \item {\bf The sims} (2000)
      \item The rise of MMORPG. {\bf World of Warcraft} (2004).
      \item The rise of casual games. Faster internet connection and free games making money on advertisement. Easy to learn and short game play. {\bf Bejeweled} (2001).
      \item The rise of social networking and games. {\bf Farmville}. 
      \item {\bf Game Boy Advance} (2001) and {\bf Game Boy Advance SP} and {\bf Nintendo DS} with double screen, and {\bf PlayStation Portable (PSP)}, and {\bf PSP Go} with downloadable content (App store approach).
      \item People started playing games on mobile devices. The release of Iphone in 2007 had a huge effect on mobile gaming. Touch screen and accelerometer raised new game types. Revenue to developers, individual small teams could make games. Not dependent on big publishers. 
      \item {\bf Xbox 360} (2005) and {\bf PlayStation 3} (2007)
      \item Xbox with online achievement system {\bf Gamescore and ranking}.
      \item {\bf Nintendo Wii} (2006) took a different direction. Weak machine, but the controllers was reveloutionary buy registering movements and the machine was cheap. They later releaset {\bf Wii fit}. 
    \end{itemize}

  \subsection{2010-2011}
    \begin{itemize}
      \item Microsoft and Sony responded to the success of Wii. Sony introduced {\bf Kinect}
      \item The big companies have big launch costs because they sell the consoles lower than the production cost, making a lot of time for recovering. They are looking at {\bf cloud} for future gaming. The broadband are not capale for this at this point. 
      \item Mobile gaming is getting increased attention. 
      \item The new hype is tablets. 
      \item {\bf Stereostopic 3D}. Modern televisions are now capable of displaying 3D. The use of glasses are not feasible. 
      \item Games are now played by different genders and ages, not just by the nerds. 
    \end{itemize}

  \section*{Example: Tennis}

    \begin{itemize}
      \item Graphics
      \item Opponents AI
      \item Storyline and setting
      \item Internet connectivity
      \item Interface and control
    \end{itemize}

  \section*{Changes in Graphics}
    \begin{itemize}
      \item Probably the most dominant change in games over the past 50 years is the change in graphics. 
      \item Sprites are small bitmaps that is initially the standard way to draw objects on the screen.
      \item Consoles had special hardware to quickly draw sprites. 
      \item Parallax scrolling was a trick in simulating 3D. images in the background moved slower than the images in the foreground. 
      \item Isometric projections. look at the game from a andgle of 45 degrees. 
      \item Graphics looked nice, but they didnt spend a lot of money on AI. When the player sees very realistic graphics he/she expects also very realistic behaviour. 
    \end{itemize}

  \section*{Changes in interaction devices}

  \section{Changes in demographics}

  \section*{Changes in gameplay}
    \begin{itemize}
      \item Adrenaline rush
      \item Highscore list
      \item difficulty level. Gradually get harder.
      \item The internet. 
    \end{itemize}

  \section*{Changes in business}
    \begin{itemize}
      \item Game design is becomming a art, and are getting more complex
      \item Artists, music, pictures, graphics, sound, effects, history, programmers, marketing, etc. 
      \item Growth in games for smartphones and other mobile devices. 
      \item Business models: advertisements, buying benefits with money.  
    \end{itemize}

  % Hardware vs. Game uniqness

### 2014-11-20 papers/paper4.tex
Added paper4

\chapter*{An Evaluation of a Mobile Game Concept for Lectures}

  \begin{itemize}
    \item Evaluation of "Lecture Quiz"
    \item Used in lectures for higher education to promote strong student participation and variation in how lectures are taught.
    \item Multiplayer quiz game
    \item Useful for testing the knowledge level and rehearsing theory
    \item Evaluated in a architecture class with students answering a questionnaire
    \item Focus was usability and usefulness
    \item The results showed that the Lecture Quiz was easy to use and contributed to increased learning. The Lecture quiz was percived as entertaining and half the students claimed they would attend more lectures if such system were used regulary. 
    \item Video games can be used insted of traditional exercises. This approach should motivate the students to put extra effort into the exercises and gives the teachers an oppertunity to monitor the students while they are doing the exercises. Second, video games can be used within lectures to improve the participation and motivation of students. 
    \item Initial main goal of the Lecture Quiz was to develop a game concept that could be used in any course that would: 
      \begin{itemize}
          \item test, motivate and engage the students
          \item The game should be using existing equipment and infrastructure available in the lecture halls at the university
      \end{itemize} 
    \item The game concept was a variant of the Sony PS2 quiz-game Buzz!
    \item Studies shows that wireless technology used in educational setting can increase mental activity, facilitate interactivity, and promote social interaction. 
    \item Characteristics that makes game fun to learn: 
      \begin{itemize}
          \item should promote the approperiate level of challenge
          \item should use fantasy and abstractions to make it more interesting
          \item trigger the players curiosity
      \end{itemize}
      The game supported curiosity, provide challenge, but not fantasy. The lack of fantasy can be compensated by making a multipplayer game were the social interaction becomes an important motivating factor. The level of challange cannot be be tailored to the individual differences.
    \item Two game modes: score distribution and last man standing. These could be used alone or together.
    \item Evaluation of the Lecture Quiz:
      \begin{itemize}
        \item Measured usability, suitability, and usefullness.
        \item Usability: satisfaction, effectiveness, and efficiency
        \item SUS score of 74.25\%. 
        \item Students find the lecture quiz engaging, students learned more, and they would be likely to attend more lectures if such games were used regularly.
        \item It is hard to measure the learning rate, but there was a significant improvement in round number two.
        \item They did not find the game distracting
      \end{itemize}
    \item Experiences from the teachers perspective
      \begin{itemize}
        \item Easy to integrate in the lecture
        \item The students had to install the clients beforehand
        \item No unnecessary delays during the lecture 
        \item "Who wants to be a millionaire"
      \end{itemize}
    \item Conclusion: The main benefit from using such game in lectures is to provide a fun way for students to review topics and to provide useful feedback to the teacher of how much the students have learned.
  \end{itemize}

### 2014-11-20 papers/paper5.tex
Added paper 5

\chapter*{What makes things fun to learn? Heuristics for Designing Instructional Computer Games}

  \section*{Introduction}
  \begin{itemize}
    \item What makes computer games fun?
    \item This paper provides a set of heuristics or guidelines for design of instructional computer games.
    \item Focus on what makes educational games fun, not what makes them educational.
    \item Essential Characteristics of good computer games:
        \begin{itemize}
            \item Challenge
            \item Fantasy
            \item Curiosity 
        \end{itemize} 
  \end{itemize}

  \clearpage
  \section*{Characteristic 1: Challenge}
    \begin{itemize}
      \item In order for a computer game to be challengeing, it must provide a goal whose attainment is uncertain. A number of important consequences follow up from this princple:
      \item {\bf Goal:} A single feature of games that correlated most strongly with preference was wether or not the game had a goal.
        \begin{itemize}
          \item {\bf Not all goals are equally good:}
            \begin{enumerate}
              \item uing the skill being taught was a means to achieving the goal, but was not the goal itself.
              \item The goal was part of an intrinsic fantasy
              \item Because of the child hero, the goal was presumably one with which the child reader could identify.
            \end{enumerate}
          \item {\bf Rich, responsive environments may be unappealing unless they provide an approperiate goal.}
            \begin{itemize}
              \item Simple games should provide an obvious goal. Goals can be made obvious or compelling by the use of visual effects or fantasy.
              \item A complex environment without built-in goals should be structured so that users will be able to easily generate goals for appriperiate difficulty. 
              \item The best goals are often practical or fantasy goals (like reaching the moon in a rocket), rather then simply goals of using a skill (like doing arithmetric problems).
              \item The players must be able to tell wether they are getting closer to a goal (performance feedback and informative feedback).
            \end{itemize}
        \end{itemize}

        \item {\bf Uncertain outcome:}
          \begin{itemize}
            \item A game is usually boring if the player is either certain to win or certain to lose. 
            \item There are four ways to make outcome of a game uncertain:
            \begin{enumerate}
              \item {\bf Variable difficulty level:} determined automatically, chosen by players, determined by the opponents skill.
              \item {\bf Multiple level goals:} score-keeping, speed responses.
              \item {\bf Hidden information}
              \item {\bf Randomness}
            \end{enumerate}
          \end{itemize}
        \item {\bf Self-esteem:}
          \begin{itemize}
            \item Success in a computer game, like success in any challenging activity, can make people feel better about themselves. 
            \item The opposite side of this principle is of cource failure in a challenging activity. Too much failure can lower a person self-esteem and can decrease a persons desire to play the game again.
            \item Performance measure feedback should be presented in a way that minimize the possibility of self-esteem damage. 
            \item There needs to be a balance between the success and failure: aslo called an approperiate difficulty level.
            \item 
          \end{itemize}
      \end{itemize}

  \section*{Characteristic 2: Fantasy}
    \begin{itemize}
      \item Games that include fantasy show or evoke images of physical objects or social situations not actually present.
      \item Non-fantasy games involve only abstract symbols
      \item {\bf Intrinsic and extrinsic fantasies:}
        \begin{itemize}
          \item One relatively easy way to try to increase the fun of learning is to take an existing curriculum and overlay it with a game in which the player progresses towards some fantasy goal, or avoids some fantasy catastophe, depending only on whether the players answers are right or wrong. 
          \item In feks hangman, the fantasy depends on the use of the skill, but not vice versa. This is called a {\bf Extrinsic fantasy}. In contrast, with {\bf intrinsic fantasy} were the skill also depends on the fantasy. 
          \item Most {\bf extrinsic fantasies} depend only on whether or not the skill is used correctly.  
          \item In {\bf intrinsic fantasies}, not only does the fantasy depend on the skill, but the skill also depends on the fantasy. This usually means that problems are presented in terms of elements of the fantasy world (Feks Dart and balloons). In intrinsic fantasy, the events in the fantasy would usually depend not just on whether the skill is used correctly, but on how its use is different from the correct use. Feks in Darts, the players can see graphically wheter their answears are too high or too low and if so by how much.
          \item In games were players are able to exploit analogies between their existing knowledge about the fantasy world and the unfamiliar things they are learning. 
          \item One potential problem with extrinsic fantasies involving a {\bf fantasy catastrophy} is that the catastrophy may be so interesting that players try to get wrong answers so they can see it. 
        \end{itemize}
      \item {\bf Emotional aspects of fantasy:}
        \begin{itemize}
          \item It is very difficult to know what emotional needs people hav eand how these needs might be partially met by computer games. It seems fair to say, however, that computer games that embody emotionally-involving fantasies like war, destruction, and competition are likely to be more popular than those with less emotional fantasies. 
          \item one obvious consequence of the importance of emotional aspects of fantasies, is that different people will find different fantasies appealing. Feks games with wars and guns can affect the genders interest in the game.  
        \end{itemize}
    \end{itemize}

  \section*{Characteristic 3: Curiosity}
    \begin{itemize}
      \item Curiosity is the motivation to learn, independent of any goal-seeking or fantasy-fulfillment. 
      \item Computer games can evoke a lerners curiosity by providing environments that have an optimal level of informal complexity. In other words, the environment should neither be too complicated nor too simple with the respect to the learners existing knowledge. 
      \item In general, an optimal complex environment will be one were the learner knows enough to have expectations about what will happen, but were these expectations are sometimes unmet. 
      \item Challenge refers to what a player can do; complexity refers to what a player can understand.
      \item {\bf Sensory curiosity:}
        \begin{itemize}
          \item Sensory curiosity involves the attention attracting value of changes of patterns in the light, sound, or other sensory stimuli of an environment. 
          \item Computer games can appeal to sensory curiosity by the use of audio and visual effects: 
            \begin{enumerate}
              \item As decoration (decorative visual effects)
              \item To enhance fantacy (like circus music to be set in the environment)
              \item As reward (when the player scores)
              \item As a representation system (to represent information more effectively than with words an numbers)
            \end{enumerate}
        \end{itemize}
      \item {\bf Cognitive curiosity:}
        \begin{itemize}
          \item Cognitive curiosity can be thought of as a desire to bring better "form" to ones knowledge structures.
          \item A way of engaging learners curiosity is to present just enough information to make their existing knowledge seem incomplete and inconsistent. The learners are then motivated to more in the order to make their cognitive structures better-formed. Feks when you read a murder mystery beside the last chapter, you have an idea of who the murder is, but you have a cognitive motivation to bring completeness to your knowledge structure by finding out who the murder was. 
          \item For instance, students may know that flowers need sun to grow, but they might not know that some plants can live in the dark. 
        \end{itemize}
        \item {\bf Informative feedback:} One way of making environments interestingly complex is to make them responsive.
          \begin{itemize}
            \item  To engage a learners curiosity, feedback should be surprising. The easy way of doing this is by using randomness. 
            \item To be educational, feedback should be constructive.  
          \end{itemize}
        \item One extreamly powerful tool for tailoring feedback to specific learners and thus maximizing their curiosity is to maintain online cognitive models of the users. 
    \end{itemize}

### 2014-11-22 papers/paper11.tex
Added paper 11

\chapter*{Scripting vs. Emergence}

\section*{Introduction}

  \begin{itemize}
    \item Scripting and emergence is two approaches to designing game environments.
    \item {\bf Scripting:} the majority of current games are implemented with scripting. Game developer predefines specific paths and interactions that the player will thake throughout the game. Relies on a given designers ideas of what is consistent and fun. Environment, objects and agents in scripted games are limited to narrow and static behaviour that the developer has predefined. 
    \item {\bf Emergence:} Is a general rule-based system that allow the creation of gameplay out of combinations of existing game elements with globally defined, consistent characteristics and behaviour. A globally designed game system provides rules and bounderies for player interaction, rather than prescripted paths. 
  \end{itemize}

\section*{Considerations for game developers}

  \subsubsection*{Developers considerations for scripted systems}
    \begin{enumerate}
      \item {\bf Effort in designing, implementing, and testing:} possible courses pf action that a player can take needs to be manually set up by the developers.
      \item {\bf Effort in modifying and extending:} scripted games scale poorly. For a bullet to break a window, there needs to be a direct relationship between the gun entity and the window entity. Fixing a bug in the system requires each instance of a game element to be visited and reconfigured manually. 
      \item {\bf Level of creative control for game developers:} The designers have full control over the game!
      \item {\bf Uncertainty and quality assurance:} Nothing occurs that was not intended or planned by the game developer. There is no uncertainty or unexpected events in the game. 
      \item {\bf Ease of feedback and direction to players:} feedback and directions to players is simple in scripted systems. As the desired outcome is known, it is straightforward to give players feedback on their success at performing actions or fulfilling goals. 
    \end{enumerate}

  \subsubsection*{Developers considerations for emergent systems}
    \begin{enumerate}
      \item {\bf Effort in designing, implementing, and testing:} creating emergent games involves designing types of objects and interactions, rather than specific ones. Rather than having a specific gun able to break a specific window, there is an additional layer of abstraction that allows a gun to break anything made of glass. (set of rules describing relationship between entities).
      \item {\bf Effort in modifying and extending:} designes scales well and is easily extended. Changes are made to objects rather than each particular instance of an object that needs to be changed. 
      \item {\bf Level of creative control for game developers:} loss of creative control for the game designer. Controlling the flow of game and telling a specific story is not as straightforward in an emergent system. 
      \item {\bf Uncertainty and quality assurance:} emergence also introduce uncertainty, making game behaviour possible that developers have never thought about.
      \item {\bf Ease of feedback and direction to players:} players have a greater need for feedback on the outcome and success of their actions in emergent systems, as the openness of the game world gives rise to more possibilities for action. 
    \end{enumerate}

\section*{Considerations for game players}

  \subsubsection*{Consistency and Immersion}

    Inconsistencies in games remind the player that it is just a game, introducing disbelief. The viewer of the player of the game is transported back to the real world, reminded disappointed that their experience was fake. Emergent systems have the potential to be used to create more consistent game worlds. For example, the player can assume that a bullet can affect anything that is damageable.

  \subsubsection*{Intuitiveness and Learning}

    Sometimes the character can climb on objects, sometimes not. Glass breaks sometimes, but sometimes they dont. It is necessary to relearn the physics of the world like a child. The intuitiveness of interactions in the game world can be partly attributed to how the interactions correspond to the interactions with the same objects in the real world. Unintuitiveness can confuse and frustrate the player. It is easier to create objects that behave and interact in more natural ways in emergent games.

    An important benefit of making game worlds more intuitive is that they become easier to learn. When learning curve is decreased, the player can spend more time playing and using less time learning.

  \subsubsection*{Emerget gameplay and player expression}

    Degree of freedom of player expression.
    When making scripted game worlds, the game is played in the exact way it was specified, which might not accomodate player creativity. Emergent gameplay allows players to solve game problems by using own strategies that was not made by the designers.

    The major difference between scripting and emergence is that emergence focuses on what the player wants to do, wheras scripting focuses on what the designer wants the player to do.

\section*{Techniques for scripting and emergence in games}

  \subsubsection*{Techniques for scriting game worlds}

    \begin{itemize}
      \item {\bf Finite State Machines (FSM):}
      \item {\bf Scripting Languages}
    \end{itemize}

  \subsubsection*{Techniques for emergence in game worlds}

    \begin{itemize}
      \item {\bf Flocking:} Is a technique for simulating natural behaviours for a group of entities. There are three different forms of steering behaviour: Separation, cohesion, alignment. 
      \item {\bf Cellular Automata:}
      \item {\bf Neural Networks:} are a machine learning technique inspired by the human brain. Can use a training set for building a desired behaviour. Can either train online of offline. 
      \item {\bf Evolutionary Algorithms:} Evolves a solution to a problem in a similar way to natural selection and evaluation. 
    \end{itemize}

### 2014-11-22 papers/paper2.tex
Added paper 11

\chapter*{Massivley Multiplayer Online Role-Playing Games: The Past, Present, and Future}

  \begin{itemize}
    \item {\BF MMORPG:} Massivley Multiplayer Online Role-Playing Games
    \item 
  \end{itemize}

### 2014-11-23 papers/paper10.tex
Added paper 9 and 10

\chapter{Evaluation of Object-Oriented Design Patterns in Game Development}

  \section*{Introduction}
    The game industry has been considered to produce revenue greater than the movie industry and its development rate has been one of the most fast growing in the USs economy. Furthermore, game design and the methods used for easier and more efficient development constitute a very interesting open research field. Futhermore, games are the first and somtimes the only market for advanced graphics techniques to demonstrate the quality of graphics they produce. The distinction between games and other forms of software is that, in games, the development groups consist of people with different field of expertise (scripters, designers, programmers, musicians, graphic artists and testers). A speculation about the absence of scientific reaserch is that games are widley considered a ``soft skill'' topic. During the past years this have changes. The use of games within education and other areas have increased it academic attention.

  \section*{Game Architecture}

    Maintainability and code reuseability are known OOP concepts, but are ideas that are extremely immature in game programming. Software is usually divided into subprograms, refered to as modules. Decomposing software into modules is an important decision that plays a main role in the architecture and further design of the program.

    There are some game modules that are essential to all games: Game logic, Level data, event handler. Other modules found in more complex games are: input, audio, graphics and dynamics.

  \section*{Design Patterns}

    ``Identified solutions to common design problems''

    If design patterns are applied properly they increase the flexability and reuseablity of the underlying system (ease future changes).
    In literature, it is not easy to find catalogues with patterns that could be used as a ``common solutions for common problems'' in games. Such catalogues would make communication and documentation easier.

    Design patterns are used, firstly  as patterns used for describing the game mechanics (gameplay and rules) and secondly as the use of OO design patterns in programming games.

    Games story: such patterns are not related to the software architecture of code. An example is the Paper-Rock-scissors pattern and is used when there are three discrete states where A defeats B, B defeates C, and C deafeates A.

    We will look further into four design patterns: Strategy, observer, State, and Bridge pattern.

    \subsubsection*{Object-Oriented design patterns in GAME LOGIC}

      {\bf Strategy pattern:} the strategy pattern, defines a family of algorithms, encapsulates each one, and makes them interchangeable. Strategy seems a very efficient way to design the way a chess game could simulate different artificial intelligence players behaviour (using different chess computers with different chess heuristics.)

      \begin{figure}[H]
        \includegraphics[width=\textwidth]{pics/strategypattern.png}
      \end{figure}

      {\bf Observer Pattern:} the observer pattern defines a one-to-many dependency between objects that when one objects changes state, all its dependants are notified and updated automatically. This can be used in a situation where a football manager game a team hires a trainer that upgrade players attributes.

      \begin{figure}[H]
        \includegraphics[width=\textwidth]{pics/observerpattern.png}
      \end{figure}

    \subsubsection*{Object-Oriented design patterns in GAME GRAPHICS}

    {\bf State Pattern:} the state pattern allows an object to alter it behaviour when its internal state change. This pattern can be used in games where an object changes the level of detail (LOD) of it apperance with respect of its distance from the camera. Feks a first person shooter, all objects of a scene should look more elegant as the main character reaches them.

      \begin{figure}[H]
        \includegraphics[width=\textwidth]{pics/statepattern.png}
      \end{figure}

    {\bf Bridge Pattern:} the bridge pattern decpuples an abstraction from its implementation so that the two can vary independently. This pattern can be used in any game with 3D objects that can be painted multiple ways. At the same time the 3D object might also vary since it can be a primitive (sphere, cube, etc) or an export from a 3D package.

      \begin{figure}[H]
        \includegraphics[width=\textwidth]{pics/bridgepattern.png}
      \end{figure}

  \section*{Evaluation}

    \subsubsection*{Software quality metrics}
    The software metrics that have been calculated can be divided into four main categories: size, complexity, coupling, and cohesion. 
    Coupling reduces the dependencies in calasses.

      \begin{itemize}
        \item Lines of Code (LOC)
        \item Number of classes (NOC)
        \item Attribute Complexity (AC)
        \item Weighted Methods per Class 1 and 2 (WMPC1 and WMPC2)
        \item Cyclomatic Complexity (CC), number of possible paths.
        \item Coupling Factors (CF)
        \item Lack of Cohesion of Methods (LCOM)
      \end{itemize}

    \subsubsection*{Evaluation Example 1 - Cannon Smash}
    The drawback in the presented approach is that the incoming data could be recieved in three different types and according to the current type a different reading strategy should be used. In the first version the programmers used a switch statement to implement this mechanism. This code suffers from needless repitation. This code is not reuseable, because the program is not flexible enough to easily adapt the extra strategies.

    A possible solution would be to use the strategy patten and in this way the complexity of the model will be decreased wheras the flexibility will increase. Attribute Complexity (AC) increased, Coupling decreased. Maintainable code.

    \subsubsection*{Evaluation Example 2 - Ice Hockey Manager}
    Ice Hockey Manager is a simulation of coaching a hockey team. In the first version they used 8 design patterns. In a later version 26 design patterns was implemented, where bridge and state was two of the patterns used.

    The state pattern represent the fact that a hockey player can either be a goalkeeper or a field player. This way, ant client of the program can create or handle a player without having to know in prior whether he is a goalkeeper or a field player.

    The bridge pattern is used in order to provide a link between player an the player attributes interfaces so that the implementation of class player can be configured at run-time.

### 2014-11-23 papers/paper3.tex
Added templates for the rest of the papers

\chapter*{Persuasive Games: Bringing Computer Entertainment Back to the Real World}

### 2014-11-23 papers/paper6.tex
Added templates for the rest of the papers

\chapter*{GameFlow: A Model for Evaluating Player Enjoyment in Games}

### 2014-11-23 papers/paper7.tex
Added templates for the rest of the papers

\chapter*{Game Development - Harder Than You Think}

### 2014-11-23 papers/paper8.tex
Added templates for the rest of the papers

\chapter*{From Visual Simulation to Virtual Reality to Games}

### 2014-11-24 papers/paper7.tex

@@
 \chapter*{Game Development - Harder Than You Think}
+
+  \section*{Introduction}
+  The hardest part of making a game has always been the engineering.
+In past times, game engineering was mainly about low-level optimization - writing code that would run quickly on the target computer, leveraging clever tricks whenever possible.
+But in the past ten years, games have ballooned in complexity.
+
+  \section*{Project size and complexity}
+
+    \subsubsection*{Tools}
+
+    \subsubsection*{Workflow}
+
+    \subsubsection*{Multiplatform Development}
+
+    \subsubsection*{Third-Party Components}
+
+    \subsubsection*{Full-Figure Options}
+
+  \section*{Highly Domain-Specific Requirements}
+
+    \subsubsection*{Engine Code}
+
+    \subsubsection*{Crosscutting Concerns}
+
+    \subsubsection*{Depth of Simulation}
+
+    \subsubsection*{Profiling}
+
+    \subsubsection*{Risk}

### 2014-11-24 papers/paper9.tex

\chapter*{Requirements Engineering and the Creative Process in the Video Game Industry}

  \section*{Introduction}

    \begin{itemize}
      \item Requirements like fun and absorbing are not well understood from the perspective of requirements engineering. 
      \item Decrease the cost of delays caused by communication errors. Locate the causes of the most costly errors.
    \end{itemize}

  \section*{Background}

    The ability to precisely communicate and capture stakeholder wants and needs is rare. Members of video game development teams include practioners from such diverse backgrounds as art, music, graphics, human factors, psychology, computer sience, and engineering.

    \subsubsection*{Emotional Factors}
    Few researchers have investigated emotional factors in requirements engineering. Feks what makes a game fun. Emotial factors is important to make a game success and human factors needs to be investigated.

    \subsubsection*{Language and Ontology}
    The game designer works with the production team to translate the vision to requirements, usually stated in natural language complete with domain spesific terminology. A common language, ontology, or vision is often mentioned as the solution to communication issues between different stakeholders.

    \subsubsection*{Elicitation, Feedback and Emergence}
    Emergent requirements discovered during the transition from preproduction to production are a significant aspect of the creative process. Communication between preprod and prod.

  \section*{Video Games Development}
  Video games are big business. However, for every advertisement for a newley released game, the trade press reports a large number of projects that fails to reach the market. What is the cuase of these failures? The multidisciplinary nature of video game development process -  with art, sound, gameplay, control systems, human factors interacting with traditional software development creates complecities that may recommend a specialized software engineering methodology for this domain.

    \subsubsection*{Development Process}
    The game process is divided into two parts: preproduction and production. The first phase is the preproduction resulting in a {\bf Game Design Document (GDD)}. Preproduction is defining the wants and needs before meeting with the development team. The other phase is the production phase. Requirements engineering, with the assistencee of the game designers, transforming the GDD into spesifications. Once the specification is complete, a traditional software development process begins, resulting in the {\bf game artifact}.
    \subsubsection*{The Game Design Document}. Moving from preproduction to production is particularly difficult in video games.

    \subsubsection*{The Game Design Document}
    The game design document is a creative work written by the game designer. If the GDD is stricted to have a formal structure, the creative process may be reduced. The GDD often includes concept statement, tagline, game genre, the story behind the game, chacters, and charcters dialog. It will also include description of how the game is played, the look, the feel, and sound of the game, the levels or missions, the cutscenes, puzzles, animations, special effects, and other elements required. The dangers with Ad hoc approach (no documentation) is the dependence of human memory for capturing decisions and their justifications. Two sets of documents needs to be maintained, and is a big cost in time and the writing stiles of the two documents is very different.

  \section*{The Transition from Preproduction to Production}
  Requirements errors are some of the most costly to fix. Too many projects violate their preproduction phases and move straight to production. Preproduction is one of the most important phases. Game Design Documents are rarely completed because it needs to be maintained throught the course of production. With time-to-market pressures, it is easy to underestimate the importance of maintainance and it is often given low priority.

  \section*{Review of Postmortem Columns}
  Game development are a competetive business and there is not much available to read for researchers. Therefore it was used postmortems from the Game Developer Magazine. This include summaries like: explain 5 goals, features of aspects of the project describing what went well and what failed.

    \begin{enumerate}
      \item {\bf Preproduction:} issues outside the traditional software development process such as inadequate game design or inadequate storyboarding.
      \item{\bf Internal:} issues related to project management and personnel.
      \item {\bf External:} issues outside the control of the development team such as changes in the market or financial conditions.
      \item {\bf Technology:} issues related to the creation or adoption of new technologies. 
      \item {\bf Schedule:} issues related to time estimates and overruns. (a part of a internal issue).
    \end{enumerate}

  Internal factor is a dominating issue in game development. This is often related to the project management issue. ``Inadequate planning'', ``underestimating the scope of tasks'', ``aggressive schedule'', ``clear goals are great when they are realastic'', ``an unrealistic schedule cant be saved without pain''.

  \section*{Examples from real games}
  The resuls from the analysis of the postmortem columns led to the conclusion that weak management of the transition from preproduction to production was a source of many issues in video game development.

    \subsubsection*{Documentation Transformation}
    Creating the requirement specification often requires a priori knowledge of the available technology so that the requirements can be presented in a context. The multiple stakeholder viewpoints, synthesizing a common domain language, numerious nonfunctional requirements, and incomsistencies as the project evolves. The associated costs are significant, leading to a strong management bias toward minimizing the documentation effort.

    \subsubsection*{Implication}
    By its nature as a creative work, a game design document is full of
    implied information. Identify these implications requires careful analysis, understanding the ramifications (a consequence of an action or event) of the implications require significant domain knowledge.

    \begin{enumerate}
      \item {\bf First level of implications:}
      \item {\bf Second level of implications:}
      \item {\bf Third level of implications:}
    \end{enumerate}

    A question is raised by identifying these three levels of implication: it is more approperiate to follow a traditional iterative process and allow these issues to surface later, or should this feedback be applied as early as possible in the process? Intuitively, eraly feedback is better. However, early feedback could have a negative effect on the creative process of the game designers.

    \subsubsection*{A Priori Knowledge}
    Domain specific terms, particularly abbrevations and acronyms, are common in working papers.
    One approach is to add professional technical writing resources to the projects. Maintaining an iteration history of sketches, such as the working paper, is challenging. Changes and justifications is an important piece of information later in the developement cycle. These justifications could lead to evolutionary chanegs in the game engine and architecture. 
    In the example, the suggested design was not supported by the technology. The success was achieved through dialog between team members, but the revised requirements and specifications for the final product was never fomally captured.

    \subsubsection*{Evaluation}
    The multiple design interations due to the technical issues cuased problems with the schedule because of the time used.

  \section*{Summary and Conclusion}
  Model for video game development that integrates preproduction with production. Project management issues are the greatest contributors to success or failure. In the case of failure, many of these issues can be tracked back to inadequate requirements engineering during the transition from preproduction to production. The introduction of production personnel into the preproduction process may have a negative effect on the creativity of the preproduction team.

  \section*{Challenges for Requirements Engineering}

  \begin{enumerate}
    \item Communication between stakeholders of different background
    \item Remaining focused on the goal and resisting feature creep
    \item Influence of prior work
    \item Media and technology interaction and integration
    \item The importance of nonfunctional requirements
    \item Gameplay requirements

  \end{enumerate}

    \subsubsection*{Meadia and Technology}
    Creating video game require the creation of numerious software artifacts.

    \subsubsection*{The Importance of NFRs}
    Video games are designed to entertain. Therefore, nonfunctional requirements such as fun, storyline, aesthetics, and flow must dominate their requirements sprecification. However, there are no established practices for capturing and specifying such NFRs. 
    Validation of gaming NFRs is very complex. The link between NFRs and target markets or user demographics has not yet been explored by RE in this domain.

    \subsubsection*{Gameplay}

### 2014-11-24 papers/paper2.tex
changed chapter headers and added content to paper7

@@
-\chapter*{Massivley Multiplayer Online Role-Playing Games: The Past, Present, and Future}
+\chapter*{2) Massivley Multiplayer Online Role-Playing Games: The Past, Present, and Future}
 
   \begin{itemize}
-    \item {\BF MMORPG:} Massivley Multiplayer Online Role-Playing Games
-    \item 
+    \item {\bf MMORPG:} Massivley Multiplayer Online Role-Playing Games
   \end{itemize}

### 2014-11-24 papers/paper7.tex
changed chapter headers and added content to paper7

@@
-\chapter*{Game Development - Harder Than You Think}
+\chapter*{7) Game Development - Harder Than You Think}
 
@@
   \section*{Project size and complexity}
+  Game development is getting more complex since the first games was published.
+The figures illustrates the expanding of nodes, providing a essential subcomponent of a game.
+Each node is a complex system.
 
     \subsubsection*{Tools}
+    To tackle the complexity, it helps to have excellent development tools.
+Sadly, this is not the case.
+On PC, the compiler development environment like Microsoft Visual Studio is basicly just a wrapper around the C++ compiler.
+most games are primarly written in C++.
+The Microsoft Visual Studio is not the best fit, but Visual C++ is the best compiler we have on PCs with no competitive alternatives, so we are just sort of along for the ride.
+
+    For the console, the life cycle is long and there isnt much motivation for for the tool-maker to improve the products toward the end of the life cycle.
+Typically, a console developer will be using an environment with only one to four years of maturity.
+
+    For simulating 3D, the program used was not created for people who make non-realtime animations, so they present a poor fit.
+Game engines are often big and slow, and they represent somthing very different from what we really need.
+    Sometimes, for building the geometry of the world, we just write our own domain-specific editors from scratch.
+
+    Historically, the situation with regard to asset management tools has also been poor.
 
     \subsubsection*{Workflow}
+    It is not only a problem with the tools, but the compile/edit/debuf time is far too long.
+Many games take half an hour or longer to compile when starting from scratch or when a major C++ header file is changed.
+
+    Once the build time have grown too long, a team may end up putting a significant amount of work into refactoring their source code to make it build more quickly.
+Often this happens too late, as the spaghetti of file dependencies has become to svere that fully refactoring it would be akin to reconstructuring the project from scratch.
+
+    Once the game is compiled, we must run it and test our changes.
+However, startup times can be very long, since games often need to load large amounts of data.
+This can cause minutes from a small code is added until it builded and up and running for testing.
 
     \subsubsection*{Multiplatform Development}
+    Many games are developed to run on multiple systems.
+During development we often have to build the game for all build types for all target platforms (PC, PSP, etc) before committing code to source control.
+Whenever this is not done, it can cause compile-time or runtime error, disrupting the work of the rest of the programming team - a bad situation.
+Before a programmer can check in a batch of changes, they may need to perform between two or five full recompiles (which mentioned earlier take a looong time!).
+The programmer can easily be waiting for hours, so theres a strong motivation to check in code as infrequently as possible.
+But they cant wait too far out of sync from the official version, causing headaches when it comes time to merge.
+A puild master can be the soulution.
+A programmer cant just sit down and get work done; there are significant barriers to push through.
 
     \subsubsection*{Third-Party Components}
+    It is desireable to be able to leverage third-party products for some of the submodules in a game in order to reduce the workload.
 

### 2014-11-25 papers/paper6.tex
changed chapter headers and added content to paper7
Added content

@@
-\chapter*{GameFlow: A Model for Evaluating Player Enjoyment in Games}
+\chapter*{6) GameFlow: A Model for Evaluating Player Enjoyment in Games}
+
+  \section*{Enjoymen in Media and Games}
+  Player enjoyment is the single most important goal for computer games.
+If Players do not enjoy the game, they will not play the game.
+Research is often focusing on usability (interface, mechanics and gameplay), rather than enjoyment in games.
+Many different models have been developed to explain and analyze media enjoyment:
+
+    \begin{itemize}
+      \item {\bf Attitude} is a psychological tendency expressed by evaluating a particular entity with some degree of favor or difavor.
+      \item {\bf Parasocial interaction} takes place when an audience member develops a parasocial relationship with a media character by talking to the character, imagining, or discussing the life of the character.
+      \item {\bf Transportation theory} suggests that the experience of enjoyment is heighted by an immersion in a narrative world, as well as from the consequence of that immersion.
+Transportation is a melding of attention, imagery, and feelings.
+      \item {\bf The theory of disposition} reletes positive or negative attitudes toward media characters to moral evaluations of their actions, as well as empathy for the main character. 
+      \item {\bf Cognition} in relation to media enjoyment involves viewers making judgements on, for instance, the characters ethics, interest, and intelligence. 
+    \end{itemize}
+
+  Each theories aims to analyse and understand the enjoyment in terms of one specific aspect or concept. Idividually these theories are fairly narrow. 
+  The aim of this study is to develop and validate a model of player enjoyment in games that is based on flow.
+
+  \section*{Enjoyment and Flow}
+  A researcher did analyse people who spend large amount of time and effort on activities that are difficult, but provide no external rewards (money, status, etc), such as chess players and rock climbers.
+Later he analysed ordinary people by asking them how it felt when their lifes were at their fullest and when what they did was most enjoyable.
+He found that optimal experience, or flow, is the same the world over.
+
+  {\it ``Flow is an experience so gratifying that people are willing to do it for its own sake, with little concern for what they will get out of it, even when it is difficult or dangerous''}.
+
+  Flow experience consists of 8 elements: 
+  \begin{enumerate}
+    \item a task can be completed;
+    \item the ability to concentrate on the task;
+    \item that concentration i possible because of the task have clear goals;
+    \item that concentration i possible because of the task provides feedback;
+    \item the ability to exercise a sense of control over actions;
+    \item a deep but effortless involement that removes awareness of the frustration of everyday life;
+    \item concern of self disappears, but sense of self emerges stronger afterwards;
+    \item the sense of the duration of time is altered
+  \end{enumerate}
+
+  \section*{Adapting Flow to Games}
+
+    \subsubsection*{Concentration}
+    To be enjoyable, a game has to require concentration, and the player must be able to concentrate on the game.
+The more concentration a task requires in terms of attention and workload, the more absorbing it will be.
+Games should grab the players attention quickly and maintain it throughout the game.
+It is important to increase the players workload but maintain an approperiate level for the players perceptual, cognitive, and memory limits.
+Also, the players shouldnt be burdened with tasks that dont feel important.
+
+    \subsubsection*{Challenge}
+    Challenge is consistently identified as the most important aspect of good game design.
+Games shouldnt be sufficiently challenging, match the players skill level, vary the level of difficulty, and keep approperiate pace (Not discouraginly hard or boringly easy).
+Satisfaction in games comes from accomplishing difficult tasks, challenging and surpassing opponents, testing skills, mastering skills, reaching a desired goal, and coping with danger.
+(the feeling of triumph).
+
+    \subsubsection*{Player Skills}
+    For games to be enjoyable, they must support player skill development and mastery.
+It is necessary that player develops their skills at playing the game.
+Intersting tutorials.
+Rewards are an important part of learning to play a game (learning rights and worngs).
+Players should also have enough information to start playing the game.
+
+    \subsubsection*{Control}
+    In order to experience flow, players must be allowed to exercise a sense of control over their actions.
+Players should be able to translate their intentions into in-game behavior and feel control of the actual movements of their character.
+    The game should be easy to use, allowing players to start the kind of game they want, turn the game on and off, and save the game in different states. 
+    Errors in the game can make players feel they have lost control, especially if the errors or consequence are out of the players control. 
+    The changes that players make in the world should be persistent, and be noticeable when players backtrack to where they were before to show that they had impact on the world. 
+    The should not be any single optimal strategy for winning and gameplays should be balanced with multiple paths throught the game or ways to win.
+
+    \subsubsection*{Clear Goals}
+    Games should present the player with a clear overriding goal early in the game. Each level should have multiple goals.
+
+    \subsubsection*{Feedback}
+    Players must recieve feeback at approperiate times.
+Feedback on direction, distance, progress towards goal.
+In-game interfaces and sound can deliver feedback.
+Games should also provide immediate feedback for players actions.
+
+    \subsubsection*{Immersion}
+    Players should experience deep but effortless involment in a game.
+Immersion, engagement, and absoption are concepts that are highlighted.
+Many game-players devoting entire nights or weekends playing games without being concurrently aware of doing so.
+Players often have a high level of emotional investment in games due to the time, effort, and attention put into playing.
+People play games to think thoughts and feel emotions that are related to work, to calm down after a hard day or to escape from everyday worries.
+Games are often seen as a form of escape from the real world or social norms, or as a way to do things that people otherwise lack the skills, resources, or social permission to.
+
+    \subsubsection*{Social Interaction}
+    Games should support and create oppertunities for social interaction.
+Social interaction is not an element of flow.
+However, it is clearly a strong element of enjoyment in games, as people play games for social interaction, whether or not they like games or the game they are player.
+To support social interaction, games should create oppertunities for player competition, cooperation, and connection.
+
+  \section*{Validating the Gameflow Critera}
+
+    \subsubsection*{Evaluation: Warcraft 3}
+
+      \begin{itemize}
+        \item {\bf Concentration:} multitude of high-quality stimuli, quickly grabbing the players attention and is held throughout the game, there is not a lot of unimportant tasks, the mission and goal feels important, and the workload is high.
+        \item {\bf Challenge:} Not met in skirmish mode since the opponent AI is hard, the difficulty is increasing, can switch race with new challenges. 
+        \item {\bf Player Skills:} Players are able to start playing Warcraft 3 immediatley, the tutorial is otional is you are an expert, new objects and features are introduced one at a time, Players are rewarded for their skill development with items, experience, and skills for their heroes.
+        \item {\bf Control:} allows players to feel in control of every aspect of the game, there is no problem with errors or recovering from errors since the game is very polished, the players have freedom and control of their actions and strategies that they use, Players can use their own strategies and choose a suited race for their strategy.
+        \item {\bf Clear Goals:} present both overriding and intermediate goals
+        \item {\bf Feedback:} Players are immediatley notified when goals or a mission objectives are completed and can check the status of their goals, subgoals, and completed goals.
+There is a notification when something new is happening.
+Players recieve their scores, broken down into heroes, units, and resources.
+        \item {\bf Immersion:} It is easy to become engrossed in Warcraft 3, as there is so much to concentrate on, many tasks to perform, things to monitor, as well as graphics, sound and animation. 
+        \item {\bf Social Interaction:} supports competition and cooperation between players, as well as social communities inside and outside the game. 
+      \end{itemize}
+
+    \subsubsection*{Evaluation: Lords of EverQuest}
+
+       \begin{itemize}
+        \item {\bf Concentration:} numerous stimuli which vary in quality.
+attractive game world, visually appealing characters, only two or three responses that becomes repetitive.
+No background story, no motivation or variation.
+The player is required to focus on unimportant tasks, and the workload is too low.
+        \item {\bf Challenge:} The challenge is below average. Challenge is poor for experienced players, Missions are straightforward, Little variation between missions, Units are very unbalanced. 
+        \item {\bf Player Skills:} ok tutorials, familiar but poor interface, Icons are hard to understand because of the details, 
+        \item {\bf Control:} do not have full control of several aspects of the game, including units, interface, and gameplay.
+One good attribute is that the interface is customizeable.
+If a player fails a mission, the player doesnt find out until the end and need to start over and replay the whole mission.
+The player does not have freedom in the strategies.
+Consequently, there is no verety in play styles or strategies available to the player.
+        \item {\bf Clear Goals:} do not present any overriding goal beside that there is a war, no goals are ever really presented. Players are told what to do, but not why. 
+        \item {\bf Feedback:} provieds sufficient feedback in general.
+The player is not given feedback on a mission until the end.
+Statistics when they level up are not visible.
+There is no difference in the terrain or visualisation of available places to build.
+        \item {\bf Immersion:} the game is too slow for players to become immersed as there is not enough to concentrate on and not enough challenge.
+Too much time spent waiting.
+There is not enough background, character development or storyline for the players to get emotionally involved in the game or connected to characters.
+        \item {\bf Social Interaction:} the game supports competition and cooperation between players. 
+      \end{itemize}
+
+    \subsubsection*{Comparisation}
+
+  \section*{Discussion and Conclusions}

### 2014-11-25 papers/paper7.tex
Added content

@@
     \subsubsection*{Third-Party Components}
-    It is desireable to be able to leverage third-party products for some of the submodules in a game in order to reduce the workload.
+    It is desireable to be able to leverage third-party products for some of the submodules in a game in order to {\bf reduce the workload}.
+Licensable third-party modules exist for some of the modules, only some of the third-party products have been successful at meeting industry needs.
+Available products covers: audio, rendering, scene management, collison detection and physics, etc.
+It can be difficult to use these third-party components; often the programmer must have a lot of experience in the problem domain in order to understand how to interface with the product successfully.
+Integrating is also a problem because of complex products with dependencies.
+Since games are CPU-intensive, it will often happen that the third-party component presents a significant {\bf performance bottleneck} for some scenarios - and the programmer must fix these situations or work around them.
+Often the third-party code fails, and a lot of work is requires, they might as well have written the module from scratch, something you dont want to find out after failing with the licensed code.
+{\bf cost/benefit analysis}.
 
     \subsubsection*{Full-Figure Options}
+    Instead of licensing components, we can license an entire game engine.
+If a company is making a game that is not doing anything new technologically, such a license can be a safe decision.
+If there is new technologically aspects, a game engine can be a poor fir as describes with third-party components.
 
   \section*{Highly Domain-Specific Requirements}
+  Currently there are three levels of programming in games: script code, gameplay code, and engine code.
+Script and gameplay code control the overall content, rules and high-level behaviour of the game.
+The engine code provides all the basic mechanisms for simulating IO.
+Engine code is more complex that gameplay code.
 
@@
 
+    When writing enige code, you need to be good at engineering and have a lot of domain-specific knowledge.
+This can be broken into two categories: {\bf Mathematical Knowledge} and {\bf algorithmic knowledge}.
+When implementing 3D there is required a lot of linear algebra and geometry (2D and 3D).
+A good engine programmer should have working familiarity with many algorithms.
+Most algorithms will be mainly focused on one task area, like rendering or physics, but these are algorithms are often very deep and take a while to master.
+
     \subsubsection*{Crosscutting Concerns}
+    To successfully build a engine, it is not enough to understand a lot of math and algorithms.
+It takes a certain experience and wisdom to choose or dicover algorithms that can be combined into a harmonious whole.
+When a engine fails, it is often because of the harmony.
+Figure 3 and 4 are complex systems with a lot of crosscutting concerns.
+Currently we do not have programming paradigms that help us address this fundamental structual problem.
 
     \subsubsection*{Depth of Simulation}
+    Game code is inherently about simulating some kind of world.
+In early games, the simulations were primitive.
+For a while we mainly focused on graphics, which is a simulation of how light behaves in the world.
+But now we are entering a time when the portions of the simulation governing physics and AI ca be more important to the end users quality and experience.
+Graphics without good AI is bad!
+When the graphics looks real, the users expects that the world behave as real as well.
+
+    {\bf Stiffness:} the fact that merely by adjusting constraints, you can cause the simulation to become unstable.
+    {\bf Tunneling:} which happens when we integrate across a timestep thats too long, causing us to miss a significant world event.
+The term tunneling comes from collision detection, where we move entities essentially by teleporting them small distances through space; if we move an entity too quickly, it may pass through a solid object like a wall, unless we take extra steps to detect that situation.
+Complex problem like $n^{2}$ do not want to be solved in real time.
 
     \subsubsection*{Profiling}
+    Were always trying to push the CPU as far as we can, so profiling is very important.
+Unfortunatley, there are no good profilers for games.
+Some vendors of graphics hardware makes some graphics-specific profiling tools, as do the makers of some game consoles.
+Those tools are also helpful but generally insufficient to get a birds eye view of the system.
 
     \subsubsection*{Risk}
+    Computer games have always evolved toward increased technical complexity to give the players things they never have experienced before.
+Game developers carry a lot of technical risks as well as game design risk.
+How will the technology interact with the system and how will the feature feel to the end user?
+Is it worth all this trouble we are taking implement it?
+
+    {\bf GAMES ARE HARD.}

### 2014-11-25 papers/paper8.tex
changed chapter headers and added content to paper7
Added content

@@
-\chapter*{From Visual Simulation to Virtual Reality to Games}
+\chapter*{8) From Visual Simulation to Virtual Reality to Games}
+
+    \section*{Defining Serious Games}
+    People respond differently to the emotionally charged term game depending on wheter they played or did not play video games while growing up.
+
+    {\bf Video Game: a mental contest, played with a computer according to certain rules for amuesements, reaction, or winning a stake.}
+
+    Developing a science of games opens a huge potential for the wider application of games in governmental or corporate training, education, health, public policy, and strategic  communication objectives.
+
+    Electronic Arts defines video games as: ``story, art, and software''. 
+    Serious games have more than just a story, art and software.
+They involve pedagogy: activities that educate or instruct.
+This addition makes games serious.
+Pedagogy must be subordinate to story - the entertainment component comes first.
+
+    \section*{Creating a Science og Games}
+    The development and wide release of the Americas Army game began an revolution in thinking about the potential role of video games for nonentertainment domains.
+A study of the game was conducted to see if  it could be used for training.
+    Example of use: when new recruits had trouble with the rifle or the obstacles course, his team had those recruits play ``Americas Army'' and required them to complete those levels in the game.
+When the reqruits did so, they then went back to the range and usually passed the range tests.
+This made us wonder what other applications would benefit from game-based instruction.
+    Mothers complained about their sons playing the game all the time and their kids knew everything about the army.
+These comments led the researchers to wonder how much of science and math education could be taught via games (learning from other than formal teaching).
+This is called ``first-person education''.
+
+    \section*{Game Production Challenges}
+    First-person shooter games are recently the only games that have reuseable engines.
+Game engines are costly to develop!
+The demand for better computer characters and story increases with the complexity of visual displays and more complex games.
+Game play innovation is becomming a competitive necessity.
+The big hits have been sports, first-person shooters, adventure, and sims-type games.
+For the game industry to continue to grow, additional genres must become more sophisticated with better backstories and better game-play technologies.
+
+    \section*{Games Reseach Agenda}
+    To influence the future of both serious and entertainment games, developers must create a research agenda that transforms the game production process form a handcrafted, labor-intensive effort to one shorter, more predictable production timelines that still manages to provide innovations and increased complexity.
+This reaseach agenda have three components: infrastructure, cognitive game design, and immersion.
+
+      \subsubsection*{Infrastructure}
+      Underlying software and hardware necessary for developing interactive games include massiveley multiplayer online game architectures, game engines and tools, streaming media, next-greneration consoles, and wireless and mobile devices.
+Game engines and tools will be vital to researchers bent on attacking the lack-of-reuse problem in gaming.
+This can also move games to the government and corporate sectors.
+Developers need an open source game engine that includes a development toolset as wideley available and utilized as Linux.
+Streaming media will play prominent role in the delivery of dynamic content to PC -based games and mobile devices.
+
+      \subsubsection*{Cognitive game design}
+      Taking a cognitive approach to game developemt will give developers the tools to create theories and methods for:
+
+        \begin{itemize}
+          \item modeling and simulating computer characters, story, and human emotion
+          \item analyzing large-scale games
+          \item innovating new game genres and play styles
+          \item integrating pedagogy with story  in the interactive game medium 
+        \end{itemize}
+
+      The Sims for a serious purpose, such as training aid for nursing.
+Developers can use this approach to model and simulate hospital operations in game form, providing an immersive experience for the nurse trainee.
+
+      It must be determined what happens during game play and how the experiences affects the players.
+The game industry has already witnessed tha failure of edutainment, an awkward combination of educational software sprinkled with game-like interfaces and cut dialog.
+This shows that story must come firstand that research must focus on combining instruction with story creation and the game development process.
+
+      \subsubsection*{Immersion}
+      Creating technologies that engage the game players mind via:
+        \begin{itemize}
+          \item computer graphics, sound, and haptics
+          \item affective computing - sensing human state and emotion
+          \item advanced user interfaces
+        \end{itemize}
+      In the next years, low cost sensors will become available that measures the players emotional state and provides this information as input when running the game.
+Game designers must understand these things if they are to engineer and implement these capabilities so that these features behave predictabily and reabily.
+      Advanced user interfaces will become key as computing moves from the standard desktop PC to mobile platforms.
+Much can be gained by studying how the game industry has developed almost universal interfaces that let gamers transision seemlessly from one game to another.
+To make significant progress in deploying serious games, developers must understand interfaces from the game perspective.
+
+    \section*{Serious Games}
+    Applying games and simulation technology to nonentertainment domains results in serious games.
+
+    \section*{Gamepipe Laboratory}
+    The first serious games must be constructed in carefully controlled university environments.
+The pipeline laboratorys mission is to research and develop interactive game technologies that dramatically shorten the production timeline, provide supporting technologies for increasing the complexity and innovation in produced games, and enable the creation of serious and entertainment games for government and corporate sponsors.
+The GamePipe Laboratory performs research and development required for the next generation game production.
+It produces serious games and game prototypes as sponsored projects.
+It also links with, and develops educational programs in game developement and game-related research and build cross-campus interdisiplinary teams to solve next-generation game technology and development problems.

### 2014-11-25 papers/paper2.tex
Added paper 2

@@
   \begin{itemize}
-    \item {\bf MMORPG:} Massivley Multiplayer Online Role-Playing Games
+    \item 
   \end{itemize}
+
+  \section*{Introduction}
+  {\bf MMORPG:} Massivley Multiplayer Online Role-Playing Games.
+Is played in a persistent online game world that continues to function even when the player logs out.
+While a player is logged out, other players can continue to play in the virtual world.
+A player assumes the role of a fictional character and then controls that characters actions.
+
+  The key aims of the research is to determine the evolutional aspects of MMORPGs. Evaluation of current and future MMORPGs.
+
+  Current research in the area of MMORPGs fits into:
+  \begin{itemize}
+    \item The social interactions between players in MMORPGs. 
+    \item The different architectures to build MMORPGs. 
+    \item The effects of latency on MMORPGs.
+    \item Problems that plague MMORPGs. 
+  \end{itemize}
+
+  The aim is to determine what is that players liked in past MMORPGs and what they want to see in future games. If this is analysed it can reduce the risk of developing MMORPG.
+
+  \section*{MMORPG Precursors}
+  MMORPGs have evolved from two definable areas: multi-user dungeons (MUDs), and computer role-playing games (CRPGs), both single and multiplayer; they in turn borrowed concepts from pen-and-paper ``Dungeons \& Dragons'' (D\&D).
+
+    \subsubsection*{Dungeons and Dragons}
+    D\&D was built from the epic advetures from the Hobbit and Lord of the rings series.
+This is the first table-top role-playing game.
+In the game the only boundary is your imagination as you play a character.
+When players create their character they must make a number of choices that dictate the tasks they can perform, their strengths and weaknesses.
+First, the player select their race or species for their character.
+Secondly, players select one of a number of predifined classes (feks fighter or magic class).
+The dungeon master is an essential component, and is responsible for setting the scene and creating the story, and assumes responsibility for all monsters and friendly characters (NPC).
+
+    The D\&D rules have served as a framework for the developers of computer games.
+
+    \subsubsection*{Character Development Models}
+
+    {\bf Class-based system:} The player chooses an initial character class when he or she first creates a character.
+The class chosen defines stengths and weaknesses.
+The class-bases system allows developers to create a game that requires players to interact and help each other by balancing strengths and weaknesses.
+
+    {\bf Skill Points-Based System:} You can implement this in two different ways: skill points are given as the player advances instead of levels of a class.
+The other way is that the player can assign the skill points to various skills and abilites in order to increase their strengths.
+A other implementation is to gain skill points in their chosen skills by simply using those skills.
+To improve healing, the player needs to perform a lot of healing.
+    This provides a more personalized character, but it can become unbalanzed when the player learn the best combination of skills.
+
+    \subsubsection*{Multiuser Dungeons}
+    The origin of multiuser dungeons (MUDs) can be tracked back to 1978, when ``MUD1'' was developed. They created an text-based game that could de played simultaneously by multiple users.
+
+    \subsubsection*{Single Player Computer RPGS}
+    RPGs brought a lot of innovative features, many of which continue to be used in MMORPGs.
+Example is realtime play and the attempt to immerse the player into the world through the use of cultures, day and night cycles, as well as weather effects.
+All these innovations are taken for granted in todays MMORPGs, but it is interesting to see when they were developed.
+
+    \subsubsection*{Multiplayer Computer RPGS}
+    Is not calles MMORPG because they only cater to a small number of players.
+``Neverwinter Nights'' indicated that online games could be successful.
+``Diablo'' introduced randomly generated dungeons and items that extended the life of the game by allowing people to replay the game with a different characters and to experience altered dungeons and items.
+A new version of the ``Neverwinter Nights'' allowed players to create their own content.
+
+  \section*{History of MMORPGS}
+
+    \subsubsection*{First Generation MMORPGS}
+    The first generation of MMORPGs were the first graphical online-role playing games that allowed many players to simultaniously play in the same universe.
+These games generally had only text-based MUDs to look as models.
+The next step was to implement user friendly interface without text commands.
+
+    {\bf The first MMORPG:} ``Meridian 59'' is often credited as the first MMORPG, as it had many of the elements that players associate with MMORPGs, at least at a early stage.
+They used the term ``persistet world'' to describe their game.
+Meridian 59 was also the first game to have a monthly charging system (business strategy).
+Meridian 59 graphics quickly outdated, but showed other developers the promising oppertunities for MMORPGS.
+
+    {\bf Ultima Online:} The first commercially successful MMORPG.
+Was based on PvP (Player vs Player).
+It also provided ``crafting'', which allows the players to build their own weapons, armor, and other objects.
+
+    {\bf EverQuest:} A blueprint for future MMORPGs.
+It had good 3D graphics and also supported a massive community.
+EQ brought MMORPGS to the mainstream.
+EQs main drive was , and is, combat, exploration, and character developement.
+EQ focused more on being a cooperative model rather than PvP; it encouraged players to work together rather than against each other.
+EQ also introduced the idea or ``raids'', which is a very large group formed to overcome extreamly difficult encounters that require cooperation of many player simultaneously.
+
+    {\bf Custamizable Interfaces:} Asheron's Call was known for the originality of its creatures.
+Used third-party tools for supporting interface enhancements.
+This is very important because many MMORPGs gets very full of different views with maps and statistics.
+
+    \subsubsection*{Second Generation MMORPGS}
+    The second generation of MMORPGs are not very innovatime, they copied concepts that was successfully implemented in the first generation MMORPGs.
+The second-generation have improved on the graphics and interface.
+
+    {\bf Realm versus Realm Combat:} ``Dark Age of Camelot'' biggest innovation was the realm-vs-realm (RvR) combat, which allowed massive battles of human-controlled players.
+
+    {\bf Instances:} ``Anarchy Online'' created ``instanced dungeons'', more commonly refered to today as ``instances'', which are special zones that generate a new copy, or instance, for each group that enters the zone.
+This allows players or groups a private copy of the zone, and ensures that there will be no competition for resources or oppertunities to kill enemies in that zone.
+Solved the problem with people killing each other for fun.
+
+    {\bf Multiplatform Support:} ``Final Fantasy XI'' was the first successful console-based MMORPG, and the first to servers with players from both consoles and PC users.
+``FF XI'' divided its goals into ``quests'' and ``missions''.
+Quests are like ordinary jobs where the player performs a task to get a reward.
+Missions advance the storyline and the progress of the player.
+Also supported the oppertunity to change jobs allows the player to experience the game differently.
+
+    {\bf One server for all:} ``Eve Online'', a galaxy game, had a lot of empty graphics and enabled to put all players in one server (galaxy).
+
+    {\bf Player Economy and Crafting:} ``Star Wars Galaxies'' enabled the most extensive set of emotes, moods, and associated animations, which, due to the way it allowed the players to express themselves.
+A big feature of the game is that almost every item in-game is created by players.
+
+    {\bf Highly Customizable Characters:} ``City of Heroes'' is said to have the most extensive character-creation system: it is able to customize everything from facial features to the superhero's outfit.
+Such customization allows players to be visually distinguishable from one another; which many MMORPG fails at: many players with the same equipment end up looking the same.
+
+    {\bf Refining the MMORPG:} ``EverQuest II'' (EQ2) one of the biggest new features was that players could become tradesmen by spending all their time crafting items to sell to the game community rather than adventuring and leveling up a character class.
+
+    {\bf World of Warcraft:}
+
+  \section*{Results and discussion}
+
+    \subsubsection*{Respondents preferences in MMORGPs}
+    The survey aimed to discover players preferences for various aspects of all MMORGPs.
+
+      \begin{itemize}
+        \item {\it ``What are the features in MMORGPs you like the most?''} \\
+        Fantasy/medieval
+        \item {\it ``What are the features in MMORGPs you like the most?''} \\
+        Lots of class/skill options, graphics and effects, large world to explore, player vs. player, combat, sosialization.
+        \item {\it ``What are the worst issues that plague MMORPGs?''} \\
+        Exploits, cheats and Item Duping.
+Running out of content, player griefing, real-world services.
+Running out of game content was ranked as a problem, while lots of content was ranked high as a desire in MMORGPs.
+This means that players are interested in bigger game world with enough content to keep them happy for years.
+As MMORGP players are already playing these games for long periods of time, they want to see many more activities and goals within these worlds.
+High latency was alsos amnong top five.
+Other problems commented was that they felt the level/monster grind was a concern, as it became repetitive and booring to kill monsters over and over again using the same basic strategy in each battle so their character could gain experience and access to new abilities and spells.
+        \item {\it ``Which style of character do you prefer?''} \\
+        The most choosed ``skill-based'' or a combination of ``skill/class-based''. A pure class-based character are very repetive. 
+      \end{itemize}
+
+  \section*{Future MMORPGs}
+
+    \subsubsection*{Improving Existing MMORPG Features}
+
+      {\it ``If your perfect MMORGP were to be released tomorrow, what three existing features would you like to see improved?''}
+
+    \begin{itemize}
+      \item {\bf Player vs Player combat (PvP):} less restrictive, unbalanced between classes becuase is sometimes follows the ``rock-paper-scissor''pattern. The need to motivate large-scale battles. 
+      \item {\bf The Level Grind:} repetitative actions and strategies. 
+      \item {\bf Storyline, World Lore, and Immersion:} distinguish a quest from a job. Job should provide money, while quests should be story-driven.  
+      \item {\bf Graphics and Effects:} characters visual apperence and customization. 
+      \item {\bf Content and Updates:} Due to the subscription fee, players expect more from developers, especially faster response times. Bugs need to be fixed faster. 
+      \item {\bf Classes and character skills:} 
+      \item {\bf Technical Enhancements:}
+      \item {\bf Item Crafting and player economy:}
+      \item {\bf Combat and skills:}
+      \item {\bf Downtime:}
+    \end{itemize}
+
+    \subsubsection*{Adding New MMORPG Features}
+
+    \begin{itemize}
+        \item {\bf Player impact in the game world}
+        \item {\bf Player-created-and-controlled content}
+        \item {\bf Technical Enhancements}
+        \item {\bf Mini games}
+        \item {\bf Item crafting and player economy}
+        \item {\bf Player aging and death}
+        \item {\bf Dynamic environments}
+        \item {\bf Dynamic content and quests}
+        \item {\bf Realtime combat and damage}
+        \item {\bf Nonplayer characters (NPC) interaction}
+        \item {\bf Evolution}
+      \end{itemize}

### 2014-12-01 papers/paper3.tex
changed chapter headers and added content to paper7
Added notes for paper 3. FINISHED!

@@
-\chapter*{Persuasive Games: Bringing Computer Entertainment Back to the Real World}
+\chapter*{3) Pervasive Games: Bringing Computer Entertainment Back to the Real World}
+
+  \section*{Introduction}
+  Computer games have some advantages that make them more popular than traditional games.
+{\it First}, they attract people by creating the illusion of being immersed in an imaginative virtual world with computer graphics and sound.
+{\it Second}, the goals of computer games are typically more interactive than that of traditional games, which brings players a stronger desire to win the game.
+{\it Third}, computer games, usually designed with an optimal level of information complexity, can easily provoke players' curiosity.
+COnsequently, computer games intrinsically motivate players by bringing them more fantasy, challenge and curiosity, which are the three main elements contributing the fun in games.
+
+  However the development of computer games has often decreased the users' physical activities and social interactions.
+Computer games focus the users' attention mainly on the computer screen or 2D/3D virtual environments, and players are bound to using keyboard, mice, and game pads while playing, thereby constraining interaction.
+{\bf To address this problem}, there is a growing trend in todays games to bring more physical movements and social interaction into games while still utilizing the benefits of computing and graphical systems.
+Thus, the real world is coming back to computer entertainment with a {\bf new game genre}, referred to as {\bf Pervasive games}, stressing the pervasive and unbiqutious nature of these games: Pervasive games are no longer confined to the virtual domain of the computer, but integrate the physical and social aspects of the real world.
+
+  \section*{Pervasive Game Genres}
+    We will now look at different types of pervasive games setting focus on different aspect og the gaming experience.
+    It is given as an example to give the reader an idea of scope and diversity of this exciting and emerging field.
+
+    \subsection*{1) Smart Toys}
+
+    Due to their shapes or forms, toys might suggest in which ways they should be played with, but in contrast to games, they are not bound by any rules or limitations on their use.
+
+    Most current realizations include traditional physical toys equipped with simple sensing technology linked to computer logic.
+The logic reacts to the changes in the toys physical state by either playing sounds or displaying graphical information.
+
+      \subsubsection*{Zowie Playsets}
+      Consists of a physical toy with movable pieces accompanied by a CD-ROM an a serial connection to a PC.
+The toy has integrated sensors that transmit the state of the playing pieces of the computer application.
+Offers several modes of play (discovery, exploration play, hands-on active play, and problem solving play).
+
+      \subsubsection*{Story Toy}
+      A rather similar approach to the zowie playsets.
+The StoryToy is a storytelling environment that uses an audio replay engine in conjunction with a tactile user interface based on a sensor network.
+The tactile interface consists of an animal farm with a multitude of animals as actors.
+The study demonstrated that audio feedback alone already creates an enjoyable level of interactivity, in addition to the traditional free play that the toys provides.
+
+      \subsubsection*{SenToy}
+      Present an interesting affective control toy, named SenToy, used to control a synthetic character in the computer game, Fantasy A.
+SenToy is a doll wirelessly connected to a PC.
+It allows players to influence the emotions of a synthetic character in FantasyA.
+Via SenToy, by using gestures associated with anger, fear, surprise, sadness, and joy, players influence the emotions of the characters they control in the game.
+
+      Players in the game FantasyA have to master SenToy and exhibit a particular set of emotions and perform a set of actions in order to evolve in the game.
+
+      Sensors are registering hands in front of eyes, walking, acceleration.
+
+    \subsection*{2) Affective Gaming}
+    One of the games of pervasive computing is to create context-aware applications that will adapt their behavior to information collected from the environment.
+
+    The {\it who} and {\it where} of a players context has been harnessed (utnyttet)  in some location-based pervasive games, while the {\it what} and {\it when} are common elements of most traditional games.
+    Capturing {\it how} a player is feeling at any given moment ant integrating this very personal representation of context into a game is {\bf the goal of affective gaming}.
+
+    Affective computing is described as ``computing that relates to, arises from, or deliberately influences emotions'' and affective gaming aims to integrate a players emotional state into the game so that the game environment can adapt to create a magical gaming experience.
+
+    Sensing an individuals emotional state is a complex and open research problem; but sensing certain aspects of a players experience while engaged in entertainment technologies is more manageable.
+The most common approach to sensing an affective state is via sensors that measures the users changing in physiological activity (skin sensors).
+Embedding sensors in chairs or game controllers, so that users can interact very naturally with their entertainment technologies.
+Other methods is thermal cameras, voice analysis, or facial expression analysis.
+
+      \subsubsection*{Utilizing Affect in a Game Environment}
+      Once a players affective state has been sensed, it can be used to inject personality into a game environment, resulting in an environment that meaningfully responds to a players context rather than to preconceived gaming challenges.
+
+      {\bf S.M.A.R.T Braingames:} uses real video games played on a PS2 integrated with NASA technology.
+The system determines whether the user is in the desired brain state by using brain waves measured by EEG, and adjusts accordingly.
+If the user maintains the desired brain state, he or she gains full control of the game controller.
+If not, the speed and steering control decrease.
+Basically, as the player maintains focus, the game responds, and when the player loses focus, ground is lost.
+The game was used to train patients to achieve a desired brain state, and not as a source of fun itself.
+The principle used in the game can be used to develop games that dynamically responds to how the player is feeling in order to create a more engaging experience.
+
+      {\bf Physiology as direct input:} Instead of using a players context to manipulate the game environment, one could use it as a direct and natural input to a game.
+For example {\bf brainball} is a game where brain waves (from EEG) are used to alter the direction in which the physical ball rolls on a physical table.
+Players sit across each other and must relax in order to make the ball move towards the opponent.
+      {\bf AffQuake} alters game play in the popular first-person shooter game via the players galvanic skin response, sensed through metal contacts on the hands or feet.
+When a player gets ``forskrekket'', the players avatar is also ``forskrekket'' and jumps back.
+It also relates the size of the players avatar to the players arousal (``opphisselse'').
+In ``Relax To Win'', a player controls the speed of a racing dragon via galvanic skin responses.
+As a player relaxes, the dragon moves faster.
+
+    \subsection*{3) Augmented (utvidet) Tabletop Games}
+    While affecting games mainly focuses on using physiological state parameters to exchange information with virtual game elements, augmented tabletop games also integrate the states of the players as central to the game experience.
+Augmented tabletop games do not serve as an input to the virtual game logic alone, but also add the richness of the social situation to the virtual domain.
+Traditional popular tabletop games are Chess and Go.
+Their continuing success can clearly be attributed to the direct interaction and communication between the players, who sit together around the same table, facing each other at an intimate distance.
+
+    Many claims that the drawback to traditional computer games is the lack of social interaction in a face-to-face setting, which tabletop games provide.
+At the same time, computer games offer the attractions of computer technology, which tabletop games lack.
+Therefore, it is only a natural evolution to combine the benefits of computer and tabletop games into a novel type of augmented tabletop games that sets out to provide new and engaging gaming experiences.
+
+      \subsubsection*{The STARS platform:} The platform consists of a dedicated hardware setup of devices such as a public vertical displays and personal digital assistants centered on a smart interactive table.
+Physical playing pieces provide a tangible interface that feels similar to the interface of a traditional board games.
+The playing pieces are detected by an overhead camera that also determines the positions of the players by tracking their hands as they reach over the tables surface.
+Additionally, an integrated RF-ID antenna detects physical tokens placed on the table surface.
+Playing pieces remains the primary interaction device during game-play because they provide the most natural interface to a board game.
+
+      The board changes during game progress. Auto-rotation boards able all players to see the same view of the board.
+
+      {\bf KnightMage} is deals with the exploration of a dungeon filled with treasures, equipments and monsters.
+In the game players must cooperate to survive against monsters in the dungeon.
+The combination of competitive and cooperative play is also found in traditional board games like Risk.
+
+      \subsubsection*{False Prophets}
+      Is a hybrid board-video game system designed to enhance player interaction.
+Its development was motivated by the various properties of board games: board games are mobile, highly interactive, provide a nonoriented interface, and allow dynamic number of players and house rules.
+They are also limited to a fairly static environment, don't allow players to save th game state, and have simple scoring rules.
+On the other hand, computer games provide complex simulations, impartial judging, evolving environments, suspension of disbelief, and the ability to save the game sate.
+But computer games often support interaction with the system, rather than with other players.
+      The goal of developing the False Prophets hybrid game was to leverage the advantages of both of computer games and board games, encouraging interaction between players.
+
+      In False Prophets, players use tangible pieces to move around a digital board projected onto a touch-sensitive table.
+The playing pieces are equipped with a button for simple game operations, while more complex interactions and private information are managed through a hand-held computer.
+
+      \subsubsection*{Smart Jigsaw Puzzle}
+      Is a hybrid tabletop game that augments the physical pieces of a jigsaw puzzle with RFID tags.
+The underlying RFID reader technology is linked to a PC application that demonstrates a virtual representation of the jigsaw puzzles physical state.
+
+    \subsection*{4) Location-Aware Games}
+    While augmented tabletop games utilize pervasive computing technology to enrich physical game boards, another popular approach in the pervasive gaming field is to regard the whole world, the architecture we live in, as a game board.
+
+    In {\bf Pirates!} players move around in the physical domain and are presented with location-dependent games on mobile computers they carry with them.
+Each PDA resembles a pirate ship, and several locations in the physical world are associated with islands the players can visit and experience via games on their PDAs.
+In contrast to recent pervasive games played outdoor, Pirates is an indoor game with a relatively small playing field.
+
+    GPS signal reception does not work indoor, and is only moderately accurate outdoors; but it allows us to cover a large playing field.
+GPS do not support communication, so additional communication channels like wifi or 3G can be used.
+
+    Some issues with large-scale pervasive games is to deal with the uncertainty in sensing and wireless communication.
+One solution is to tailor the technical shortcomings and make them a part of the game experience.
+
+      \subsubsection*{Treasures} 
+      Is a pervasive multilayer game played on an outdoor area of several thousand square meters.
+The game revolves around collecting virtual coins that are hidden in the area.
+They players need to be at the physical location to collect the coins.
+A key aspect is the seamless integration of technology into out everyday lives.
+      There will might be areas where there is no wifi connection. This can be integrated into the game play so players can hide in the shadows of the missing connectivity.
+
+      \subsubsection*{Can you see me now?}
+      Combines pervasive gaming action in a real part of a city with online game-play in a virtual model of the game area.
+In real steets, runners are equippted with GPS and WiFi run around to catch the online players that move through the virtual representation of the streets (real runners and online players).
+The runners use walkie-talkie to coordinate their movements to catch the online players.
+The online players sit at a home computer listening to the walkietalkie via stream audio.
+Text messages can be exchanged for coordination between online players.
+
+      \subsubsection*{Uncle Roy All Around You}
+      Is a successor of ``Can You See Me Now?''.
+CYSMN managed to mix online and physical games, URAAY picks up the concept of these two domains.
+But it also integrates more aspects of the real world into the game mechanics.
+In particular, physical players remains constantly uncertain about which part of the real environment are actual parts of the game.
+For instance, through communication with online players or the game itself, players in the streets recive hints such as gender or color of other important persons in their vicinity, effectively integrating passers-by into the action.
+
+      The story is about finding, somewhere in the city, a mysterious character known as Uncle Roy, with the physical players searching in the streets.
+They are supported by online players who track their progress an aid them with hints about the way to Uncle Roy.
+
+      \subsubsection*{Catch the flag}
+      In a augmented twist on the popular traditional game, smart phones are used as th main interface, enabling physical role play, as players are not confined to desktop computers.
+Players can move about freely in the real world over wide area outdoor spaces, while maintaining seamless real-time networked contact with other players in both the real and virtual world.
+      This game is an popular outdoor game for children.
+In the mixed reality version, each team ha at least one knight and one guide, as well as other entities including bombs, traps, flags, magic potions and castles.
+Throughout the game, guides use a desktop application to assist allied knights by giving guidelines and setting magic potions, which for a short period of time can turn knight into warriors, as well as catch opponents knights by setting traps.
+Knights navigate and receive guidelines though a smartphone, and can catch the opponents flag by acquiring the Bluetooth object representing the flag in the real world.
+Any team who can safely bring the opponents flag to its base wins the game.
+
+    \subsection*{5) Augmented Reality Games}
+    Perhaps the most technically advanced pervasive games use augmented reality techniques as a bias.
+Users see their view augmented with 3D objects registered such that they appear to exist in real space.
+
+    Augmented reality can be created with different technologies:
+    Using head-mounted displays, using images projected on real world surface, or using hand-held devices.
+
+      \subsubsection*{The characteristics of AR Games}
+      In augmented reality, players are not immersed in virtual content, rather, virtual elements are added to the real world.
+It is stated that the ideal entertainment experiences comes from the combination of physical experience, virtual content, storytelling, and the imagination of the user.
+Augmented reality offers both physical and virtual aspects, leaving creative designers to simulate the imagination.
+
+      Many early augmented reality games have focused more on the technology rather then the game design.
+
+      \subsubsection*{$AR^{2}$ Hockey}
+      Players as physical paddles and the puck is virtual.
+The projects intent was to demonstrate technology, it can serve to illustrate the possibility of new game features.
+PingPongPlus is also a similar implementation.
+In it, the ball remains physical, but the system tracks it movement and presents visual effects on the game table using a projector.
+
+      \subsubsection*{ARQuake}
+      The ARQuacke projects demonstrates
+
+      \subsubsection*{Human Pacman}
+      Human players who takes the role of pacman.
+Pacman and Ghost players can see, via head-mounted displays.
+The cookies are shown from a first-person shooter perspective.
+It lacked social interaction, so a helper is introduced.
+
+      \subsubsection*{Tilt-Pad Pacman}
+      A using wearing Head-mounted display sees the 3D world and controls the pacman by tilting a heldhand device.
+
+      \subsubsection*{Magic Land}
+      Is an augmented reality environment in which 3D avatars of live human beings and 3D computer-generated virtual animations play and interact.
+The environment has two main areas: the recording room and the interactive room.
+The recording room is where users can have themselves captured and made into live 3D models that will interact in the mixed-reality scene.
+After being captured, the user can go to the interactive room to play with their own figure.
+
+      \subsubsection*{AR Worms}
+
+      \subsubsection*{AR Tankwar}
+
+      \subsubsection*{Open Issues in AR Games}
+      As well as offering a unique combination of characteristics for game design, augmented reality presents some unique problems. 
+      For example, there is limited collaboration , particularly with the use of head mounted displays.
+
+      The development of interaction metaphors is still immature (paddle, pointer, and lens-based interactions are some examples).
+While augmented reality may allow players to use their whole bodies to interact with the game, research is needed to determine what metaphors are most effective in different gaming situations.
+
+      Current augmented reality systems require especially configured hardware and controlled environments.
+Such systems are expensive to set up, and so commercial game development is not viable.
+In the near future, hand-held augmented reality games may become popular via mobile phones; but head-mounted displays may require drastic changes in technology or the marketplace to become viable.
+Fixed installations and other business models may be a option.
+
+      Augmented reality offers many possibilities for game design.
+Games designed using it may draw the characteristics of both real-world and computer games.
+However, there are significant challenges due to its expense, difficulties with hardware, and lack of design guidelines.
